Software Architectures for Integrating Workflow Engines in Science Gateways

### 2016-05-13 paper.tex
completed

\caption{Architecture evaluation. Lower values (brighter colors) indicate better performance. Cell color is set as follows: (1) on each row, metric values are
  normalized between 0 (best value) and 1 (worst value):
  $m'=\frac{m-m_{\mathrm{min}}}{m_{\mathrm{max}}-m_{\mathrm{min}}}$ where
  $m$ is the metric value, $m_{\mathrm{min}}$ and $m_{\mathrm{max}}$
  are the minimal and maximal values among all architectures; (2) the RGB hexadecimal color code of the cell
  is \texttt{\#99XX99}, where X=\texttt{round}(F-6m') (\texttt{round} rounds a number to the nearest integer). \closednote{Tristan}{Use 6. Round to the nearest integer.}{Tristan}{Fixed.}}
\label{table:evaluation}
\end{table*}

@@

The pool architecture is overall the most scalable, which is not
surprising since it is designed precisely for scalability. Tight
integration is the least scalable and all the other architectures
perform the same overall. The global scalability score, however,
should not conceal \closednote{Naj}{what do you mean by
  mask?}{Tristan}{I guess 'hide' or 'conceal' are better
  words. Fixed.} the unique characteristics of architectures regarding
this criterion. Nested workflows are the only architecture that can
easily accommodate distributed workflow execution, which can be
critical in some cases. At the same time, the scheduling constraints
created by task encapsulation and nested workflows may become problematic \closednote{Naj}{please don't use these from-film-marketing
  words?! :-) }{Tristan}{OK!} depending on the type of targeted
infrastructure. Non-reliable infrastructures \closednote{Naj}{you mean by
  the showstoppers, i.e. the nested workflows which can accommodate
  distributed executions but are limited by scheduling constraints?}{Tristan}{Yes. I rephrased a bit this part.} ,
for example, could hardly cope with workflow engines being wrapped in
computing tasks as done in task encapsulation. The availability of
multiple engine instances could also become a critical feature for
science gateways with important workloads, which would favor task
encapsulation and pool. \note{Naj}{I feel this paragraph is a bit
  vague, and doesn't distill and contextualize the thorough work
  described before.}

Differences in other specific features should not be neglected. Nested
workflows (with service invocation) and workflow import (with service
invocation) are the only architectures that support meta-workflows;
task encapsulation and nested workflows may complexify
fine-grained debugging. \closednote{Rafael}{I would suggest to provide
  some more examples when this could be interesting.}{Tristan}{Done.}
The availability of fine-grained debugging information may also be
critical to efficient user support.

Table~\ref{table:overall} provides an overall comparison between
architectures, based on the metrics in Table~\ref{table:evaluation}.
Note that this analysis is only for illustration purposes since it
assumes that all criteria have equal weights while real systems would
favor some of them. Overall, task encapsulation, and workflow import (with
service invocation) perform a bit better than tight integration,
service invocation and pool. Nested workflows stands a bit behind for
the reasons mentioned previously.\closednote{Rafael}{I would emphasize
  that this analysis is only for illustration purposes, since it does
  not account for different weights in metrics.}{Tristan}{Done.}
\begin{table*}
\footnotesize
\centering
\begin{tabular}{rcccccc}
                                    & \textbf{Tight}
                                    & \textbf{Service}
                                    & \textbf{Task}
                                    & \textbf{Pool}
                                    & \textbf{Nested}
                                    & \textbf{Import} \\
  Integration (global score)    &
                                    \cellcolor[HTML]{99FF99}{0.00}
                                    & \cellcolor[HTML]{99E899}{0.22}
                                    & \cellcolor[HTML]{99FF99}{0.00}
                                    & \cellcolor[HTML]{99D299}{0.44}
                                    & \cellcolor[HTML]{999999}{1.00}
                                      & \cellcolor[HTML]{99DD99}{0.33}\\
Robustness (global score) &
                                \cellcolor[HTML]{99F099}{0.14}
                                    & \cellcolor[HTML]{99D399}{0.43}
                                    & \cellcolor[HTML]{99FF99}{0.00}
                                    & \cellcolor[HTML]{99B699}{0.71}
                                    & \cellcolor[HTML]{999999}{1.00}
                                    & \cellcolor[HTML]{99D399}{0.43}\\
  Extensibility (global score)  & \cellcolor[HTML]{99D299}{0.44}
                                     & \cellcolor[HTML]{99D299}{0.44}
                                     & \cellcolor[HTML]{99FF99}{0.00}
                                     & \cellcolor[HTML]{99E899}{0.22}
                                     & \cellcolor[HTML]{999999}{1.00}
                                     & \cellcolor[HTML]{99E899}{0.22}\\
Scalability (global score)  & \cellcolor[HTML]{999999}{1.00}
                                     & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{99FF99}{0.00}
                                     & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{99CC99}{0.50}\\
Specific features (global score) & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{999999}{1.00}
                                     & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{99FF99}{0.00}\\
                                    & \cellcolor[HTML]{99E799}\textbf{2.1}
                                    & \cellcolor[HTML]{99E699}\textbf{2.1}
                                    & \cellcolor[HTML]{99FE99}\textbf{1.5}
                                    & \cellcolor[HTML]{99EF99}\textbf{1.9}
                                    & \cellcolor[HTML]{999999}\textbf{4.0}
                                    & \cellcolor[HTML]{99FF99}\textbf{1.5}\\
\end{tabular}
\caption{Overall evaluation. Brighter colors and lower scores indicate better performance. Scores
  are obtained by summing the normalized global scores ($m'$ values) of
  each criterion in Table~\ref{table:evaluation} and the colors are obtained as in Table~\ref{table:evaluation}. }
\label{table:overall}
\end{table*}

@@

A few limitations should be considered when using the results of our
evaluation. First, our evaluation methods aimed to provide comparative
metrics, however the high-level of abstraction used to derive these
metrics hinders the complexity of real systems to some
extent.\closednote{Naj}{Here, you were bashing your article harshly,
  so I rewrote it}{Tristan}{yup, thanks!}. In particular, the
presented architectures are abstract patterns that may be mixed
together in actual systems.\answerednote{Naj}{something about "theoretical
  pattern", and "blending" is jarring, are these common words in the
  computer sci context?}{Tristan}{Better now?} The distinction between tight integration and
service invocation, for instance, may not always be that clear in
practice. Service invocation may also be combined with task
encapsulation in some cases. However, the criteria discussed in this
work would still apply to broadly compare and categorize such cases of
hybrid architectures.

Moreover, all the interactions involved in the architectures were
treated equally while some of them are obviously more complex than
others. For example, interaction \texttt{g} (workflow language
conversion) is clearly more complex than interaction \texttt{d}
(service invocation), and may create much more reliability issues as
well. However, without entering into the details of a particular
system, quantification of the robustness or the amount of development
associated with each interaction and software component will not be
precise.

The particular case of interaction \texttt{g} (workflow language
conversion) requires particular attention when implementing a real
system \closednote{Naj}{so are you discussing it here, or in the
  future?}{Tristan}{It's out of scope, a complete paper in itself. Rephrased this part.} since this interaction may not be easily generalized to any
workflow language. For instance, converting FSL, PSOM or Nipype
pipelines to any other workflow language is problematic because these
engines rely on general-purpose programming languages such as Bash,
Octave and Python, which are much richer than scientific workflow
languages.

@@

Finally, workflow engines are only one of the many aspects involved in
the design of a science gateway. Other properties definitely influence
the design of a software architecture, for instance collaborative
features, data visualization, search, authentication, accounting and
so on.

\section{Conclusion}

@@

% Impact of the work
To the best of our knowledge, our work is the first to systematically
review and compare software architectures used to integrate workflow
engines in science gateways.\closednote{Naj}{instead of its kind,
  explain what kind it is, e.g. systematic review of all
  ...}{Tristan}{Done.} So far, the literature on science gateways and
workflow engines has focused on the description of particular systems,
or on the presentation of a particular architecture.  Instead, our
analysis abstracts and evaluates architectural patterns independently
from any particular system. It provides general insight to integrate
science gateways and workflow engines, which we hope will be useful
for software architects.

>>>>

\caption{Architecture evaluation. Lower values (brighter colors) indicate better performance. Cell color is set as follows: (1) on each row, metric values are
  normalized between 0 (best value) and 1 (worst value):
  $m'=\frac{m-m_{\mathrm{min}}}{m_{\mathrm{max}}-m_{\mathrm{min}}}$ where
  $m$ is the metric value, $m_{\mathrm{min}}$ and $m_{\mathrm{max}}$
  are the minimum and maximum values for all architectures; (2) the RGB hexadecimal color code of the cell
  is \texttt{\#99XX99}, where X=\texttt{round}(F-6m') (\texttt{round} rounds a number to the nearest integer). \closednote{Tristan}{Use 6. Round to the nearest integer.}{Tristan}{Fixed.}}
\label{table:evaluation}
\end{table*}

@@

The pool architecture is overall the most scalable, which is not
surprising since it is designed precisely for scalability. Tight
integration is the least scalable and all the other architectures
perform the same overall. The global scalability score, however,
should not conceal \closednote{Naj}{what do you mean by
  mask?}{Tristan}{I guess 'hide' or 'conceal' are better
  words. Fixed.} the unique characteristics of architectures regarding
this criterion. Nested workflows are the only architecture that can
easily accommodate distributed workflow execution, which can be
critical in some cases. At the same time, the scheduling constraints
created by task encapsulation and nested workflows may become problematic \closednote{Naj}{please don't use these from-film-marketing
  words?! :-) }{Tristan}{OK!} depending on the type of targeted
infrastructure. Non-reliable infrastructures\closednote{Naj}{you mean by
  the showstoppers, i.e. the nested workflows which can accommodate
  distributed executions but are limited by scheduling constraints?}{Tristan}{Yes. I rephrased a bit this part.} ,
for example, could hardly cope with workflow engines being wrapped in
computing tasks as done in task encapsulation. The availability of
multiple engine instances could also become a critical feature for
science gateways with important workloads, which would favor task
encapsulation and pool. \note{Naj}{I feel this paragraph is a bit
  vague, and doesn't distill and contextualize the thorough work
  described before.}

Differences in other specific features should not be neglected. Nested
workflows (with service invocation) and workflow import (with service
invocation) are the only architectures that support meta-workflows;
task encapsulation and nested workflows may hamper
fine-grained debugging. \closednote{Rafael}{I would suggest to provide
  some more examples when this could be interesting.}{Tristan}{Done.}
The availability of fine-grained debugging information may also be
critical to efficient user support.

Table~\ref{table:overall} provides an overall comparison between
architectures, based on the metrics in Table~\ref{table:evaluation}.
Note that this analysis is only meant for illustration purposes, since it
assumes that all criteria have equal weights while real systems would
favor some of them. Overall, task encapsulation, and workflow import (with
service invocation) perform a bit better than tight integration,
service invocation and pool. Nested workflows stand a bit behind for
the reasons mentioned previously.\closednote{Rafael}{I would emphasize
  that this analysis is only for illustration purposes, since it does
  not account for different weights in metrics.}{Tristan}{Done.}
  Integration (global score)    &
                                    \cellcolor[HTML]{99FF99}{0.00}
                                    & \cellcolor[HTML]{99E899}{0.22}
                                    & \cellcolor[HTML]{99FF99}{0.00}
                                    & \cellcolor[HTML]{99D299}{0.44}
                                    & \cellcolor[HTML]{999999}{1.00}
                                      & \cellcolor[HTML]{99DD99}{0.33}\\
Robustness (global score) &
                                \cellcolor[HTML]{99F099}{0.14}
                                    & \cellcolor[HTML]{99D399}{0.43}
                                    & \cellcolor[HTML]{99FF99}{0.00}
                                    & \cellcolor[HTML]{99B699}{0.71}
                                    & \cellcolor[HTML]{999999}{1.00}
                                    & \cellcolor[HTML]{99D399}{0.43}\\
  Extensibility (global score)  & \cellcolor[HTML]{99D299}{0.44}
                                     & \cellcolor[HTML]{99D299}{0.44}
                                     & \cellcolor[HTML]{99FF99}{0.00}
                                     & \cellcolor[HTML]{99E899}{0.22}
                                     & \cellcolor[HTML]{999999}{1.00}
                                     & \cellcolor[HTML]{99E899}{0.22}\\
Scalability (global score)  & \cellcolor[HTML]{999999}{1.00}
                                     & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{99FF99}{0.00}
                                     & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{99CC99}{0.50}\\
Specific features (global score) & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{999999}{1.00}
                                     & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{99CC99}{0.50}
                                     & \cellcolor[HTML]{99FF99}{0.00}\\
                                    & \cellcolor[HTML]{99E799}\textbf{2.1}
                                    & \cellcolor[HTML]{99E699}\textbf{2.1}
                                    & \cellcolor[HTML]{99FE99}\textbf{1.5}
                                    & \cellcolor[HTML]{99EF99}\textbf{1.9}
                                    & \cellcolor[HTML]{999999}\textbf{4.0}
                                    & \cellcolor[HTML]{99FF99}\textbf{1.5}\\
\caption{Overall evaluation. Brighter colors and lower scores indicate better performance. Scores
  are obtained by summing the normalized global scores ($m'$ values) of
  each criterion in Table~\ref{table:evaluation} and the colors are obtained as in Table~\ref{table:evaluation}. }
\label{table:overall}
\end{table*}

@@

A few limitations should be considered when using the results of our
evaluation. First, our evaluation methods aimed to provide comparative
metrics, however the high-level of abstraction used to derive these
metrics hinders the complexity of real systems to some
extent.\closednote{Naj}{Here, you were bashing your article harshly,
  so I rewrote it}{Tristan}{yup, thanks!} In particular, the
presented architectures are abstract patterns that may be mixed
together in actual systems.\answerednote{Naj}{something about "theoretical
  pattern", and "blending" is jarring, are these common words in the
  computer sci context?}{Tristan}{Better now?} The distinction between tight integration and
service invocation, for instance, may not always be that clear in
practice. Service invocation may also be combined with task
encapsulation in some cases. However, the criteria discussed in this
work would still apply to broadly compare and categorize such cases of
hybrid architectures.

Moreover, all the interactions involved in the architectures were
treated equally, whereas some of them are obviously more complex than
others. For example, interaction \texttt{g} (workflow language
conversion) is clearly more complex than interaction \texttt{d}
(service invocation), and may challenge reliability more strongly. However, without entering into the details of a particular
system, quantification of the robustness or the amount of development
associated with each interaction and software component will hardly be
precise.

The particular case of interaction \texttt{g} (workflow language
conversion) requires particular attention when implementing a real
system \closednote{Naj}{so are you discussing it here, or in the
  future?}{Tristan}{It's out of scope, a complete paper in itself. Rephrased this part.} since this interaction may not be easily generalizable to any
workflow language. For instance, converting FSL, PSOM or Nipype
pipelines to any other workflow language is problematic because these
engines rely on general-purpose programming languages such as Bash,
Octave and Python, which are much richer than scientific workflow
languages.

@@

Finally, workflow engines are only one of the many aspects involved in
the design of a science gateway. Other properties definitely influence
the design of a software architecture, for instance collaborative
features, data visualization, search, authentication, accounting and
so on.
\note{sil}{do you want to refer to the science gateway canvas paper here?
	@inproceedings{shahand:2015ab,
		Address = {Portland, OR, USA},
		Author = {Shahand, Shayan and van Kampen, Antoine H. C. and Olabarriaga, S\'{\i}lvia D.},
		Booktitle = {{Proceedings of the Science of Cyberinfrastructure: Research, Experience, Applications and Models}},
		Series = {SCREAM'15},
		Title = {{Science Gateway Canvas: A business reference model for Science Gateways}},
		Year = {2015},
	}
\section{Conclusion}

@@

% Impact of the work
To the best of our knowledge, our work is the first to systematically
review and compare software architectures used to integrate workflow
engines into science gateways.\closednote{Naj}{instead of its kind,
  explain what kind it is, e.g. systematic review of all
  ...}{Tristan}{Done.} So far, the literature on science gateways and
workflow engines has focused on the description of particular systems,
or on the presentation of a particular architecture.  Instead, our
analysis abstracts and evaluates architectural patterns independently
from any particular system. It provides general insight to integrate
science gateways and workflow engines, which we hope will be useful
for software architects.
